---
title: "Homework 3 - Time series"
author: "Mateus Auza Cruz"
date: today 
format: 
  pdf:
    pdf-engine: xelatex
    documentclass: scrartcl
    papersize: letter
    colorlinks: true
    fig-cap-location: bottom
    fig-pos: H
    geometry:
      - margin=1in
    fontsize: 11pt
execute:
  echo: true
  warning: false
  message: false

editor:
  markdown:
    wrap: 80

df-print: kable
---

```{r}
#| label: packages
#| message: false
#| warning: false
#| include: false

par(cex.main = 0.8)
```


```{r}
milk=read.table("milk.txt", header=F)
milk= ts(milk, start=1962, frequency=12)
```


Our time series analysis looks like this:
```{r}
plot(milk, ylab="milk", main="Monthly milk production per cow")
```

From this plot we can see that the data is non stationary and there seems to
have a seasonal and an non stable variance. To better visualize the seasonal
component (and to check that indeed we have an $S=12$) we will do the following
plot:

```{r}
#| fig-width: 12
#| fig-height: 5
plot(milk[1:12], type="l", ylim=c(min(milk), max(milk)))
for(i in 2:16){
  lines(milk[12*(i-1)+(1:12)],col=i)
}
legend("topleft", legend = 1962:1976, lty = 1, col = 1:15, cex = 0.5)
```

The plot suggests a trend and seasonality, so we consider $d=1$ and $D=1$. We also have a non stationary variance. Thus we need to
correct this. First we will correct the variance using a log transformation.

```{r}
log.milk=log(milk)
plot(log.milk, ylab="log(milk)", main="Monthly milk production per cow (log scale)")
```


Then we remove the seasonality:

```{r}
no.season=diff(log.milk,lag=12)
plot(no.season, main = "Seasonality in 'milk_log' (S = 12)", ylab="milk")
```

Now we remove the linear trend by differencing once.

```{r}
no.dep=diff(no.season, lag=1)

plot(no.dep,ylab= "milk", main= "Remove linear trend in 'milk_noseason' (d = 1)")

```

Now, we can see that this time series looks **stationary** and thus we can use it to
forecast data.

Now our goal is to fit the best model for our data, this will be made though the
stationary data. To do that, we wil plot the **acf** and the **pacf**.

```{r}
#| fig-width: 12
#| fig-height: 5
# Yearly component
par(mfrow=c(1,2))
acf(no.dep, lag.max=100, main="ACF of the year component (Q)")
pacf(no.dep, lag.max=100, main="PACF of the year component (P)")

# Monthly component
par(mfrow=c(1,2))
acf(no.dep, lag.max=12, main="ACF of the month component (q)")
pacf(no.dep, lag.max=12,main="PACF of the month component (p)")

```

From the ACF and PACF plots, we can identify possible components for a SARIMA model.

First, looking at the **ACF of the seasonal (yearly) component**, we see a clear spike at lag 12 that goes beyond the 95% confidence bounds. After that, there is only a small spike around lag 48, and the values slowly decrease. This suggests a **seasonal MA component**, with order \(Q = 0\) or \(Q = 1\), since the pattern seems to tail off.

Next, in the **PACF of the seasonal component**, we notice significant spikes at multiples of 12 (lags 12, 24, and 36), although they become weaker after lag 36. This pattern is typical of a **seasonal AR component**, so we can reasonably choose \(P = 1\).

For the **ACF of the non-seasonal (monthly) component**, there is a spike at lag 1 (besides lag 0), followed by a gradual decrease. This points to a **non-seasonal MA component** with order \(q = 0\) or \(q = 1\).

Finally, the **PACF of the non-seasonal component** shows a clear spike at lag 1 and then quickly decreases. This suggests a **non-seasonal AR component** with order \(p = 0\) or \(p = 1\).

Based on these observations, we consider the following candidate models:

- **SARIMA(1, 1, 0)(1, 1, 0)**  
- **SARIMA(0, 1, 1)(1, 1, 0)**  
- **SARIMA(1, 1, 1)(1, 1, 0)**  
- **SARIMA(1, 1, 0)(1, 1, 1)**  
- **SARIMA(0, 1, 1)(1, 1, 1)**  
- **SARIMA(1, 1, 1)(1, 1, 1)**  

We will now try to choose the best model through the AIC/BIC criteria:

```{r}
fit1= arima(milk, order= c(1,1,0), seasonal= list(order=c(1,1,0), period=12))
fit2= arima(milk, order= c(0,1,1), seasonal= list(order=c(1,1,0), period=12))
fit3= arima(milk, order= c(1,1,1), seasonal= list(order=c(1,1,0), period=12))
fit4= arima(milk, order= c(1,1,0), seasonal= list(order=c(1,1,1), period=12))
fit5= arima(milk, order= c(0,1,1), seasonal= list(order=c(1,1,1), period=12))
fit6= arima(milk, order= c(1,1,1), seasonal= list(order=c(1,1,1), period=12))

data.frame(
  Model = paste0("fit", 1:6),
  AIC = c(AIC(fit1), AIC(fit2), AIC(fit3), AIC(fit4), AIC(fit5), AIC(fit6)),
  BIC = c(BIC(fit1), BIC(fit2), BIC(fit3), BIC(fit4), BIC(fit5), BIC(fit6))
)
```
The three best models according to these criteria are:
**SARIMA(1,1,0)(1,1,1)**, **SARIMA(0,1,1)(1,1,1)**, and **SARIMA(1,1,1)(1,1,1)**.

We now assess the adequacy of these models using the Ljung–Box test, in order to check for remaining autocorrelation in the residuals.

```{r}
#| fig-width: 12
#| fig-height: 5
LjungBoxPlot <- function(res, ARord, MAord, lag.max = 20){
  pval <- numeric(lag.max)
  tot <- ARord + MAord
  for(k in (tot + 1):lag.max){
    pval[k] <- Box.test(res, lag = k, type = "Ljung", fitdf = tot)$p.value
  }

  plot((tot+1):lag.max, pval[(tot+1):lag.max],
       type="b", pch=19, ylim=c(0,1),
       xlab="Lag", ylab="p-value", main="Ljung-Box test")
  abline(h = 0.05, col="blue", lty=2)
}

resid4= residuals(fit4)
resid5= residuals(fit5)
resid6= residuals(fit6)

par(mfrow = c(3,1), mar = c(2, 6, 2, 1))
LjungBoxPlot(resid4, ARord = 2, MAord = 1)
LjungBoxPlot(resid5, ARord = 1, MAord = 2)
LjungBoxPlot(resid6, ARord = 2, MAord = 2)

```
The Ljung–Box test does not indicate significant autocorrelation in the residuals, suggesting that all three models provide an adequate fit. Among these models, **SARIMA(1,1,0)(1,1,1) (fit4)** has the lowest AIC, and is therefore selected as the final model.


Let us now estimate the parameters:
```{r}
fit4
```
The model can be written in backshift notation as
$$
(1 - 0.2276B)(1 - 0.0650B^{12})(1 - B)(1 - B^{12}) X_t
=
(1 - 0.5822B^{12}) \varepsilon_t.
$$

Expanding this expression:
$$
\begin{aligned}
X_t =\;& 1.2276\,X_{t-1} - 0.2276\,X_{t-2} \\
      &+ 1.0650\,X_{t-12} - 1.307394\,X_{t-13} + 0.242394\,X_{t-14} \\
      &- 0.0650\,X_{t-24} + 0.079794\,X_{t-25} - 0.014794\,X_{t-26} \\
      &+ \varepsilon_t - 0.5822\,\varepsilon_{t-12}.
\end{aligned}
$$

Now, that we have the equation of the model, we will now try to predict the 15 values ahead:

```{r}
#| fig-width: 13
#| fig-height: 7

pred.ts <- predict(fit4, n.ahead = 15)

ts.tot <- ts(c(milk, pred.ts$pred), start = 1962, frequency = 12)

plot(ts.tot, type = "l", col = "blue",
     main = "Forecast of the milk time series")

lines(ts(pred.ts$pred + 1.96 * pred.ts$se,
         start = tsp(milk)[2] + 1/12, frequency = 12),
      lty = 2, col = "red")

lines(ts(pred.ts$pred - 1.96 * pred.ts$se,
         start = tsp(milk)[2] + 1/12, frequency = 12),
      lty = 2, col = "red")
```








