---
title: "Arma-modeling"
author: "Mateus Auza Cruz (28732200)"
date: today

format:
  pdf:
    pdf-engine: xelatex
    documentclass: scrartcl
    papersize: letter
    colorlinks: true
    fig-cap-location: bottom
    fig-pos: H
    geometry:
      - margin=1in
    fontsize: 11pt

execute:
  echo: true
  warning: false
  message: false

editor:
  markdown:
    wrap: 80

df-print: kable
---

## Data Loading

```{r}
serie4= read.table("serie4.txt", header=T); X4= ts(serie4)
```

# Exercice 1 - Visual analysis and order estimation
```{r}
#| fig-width: 7
#| fig-height: 6
#| results: hold
par(mfrow=c(3,1)); plot(X4); acf(X4); pacf(X4)
```

From the ACF and PACF plots, neither function exhibits a clear cutoff; both gradually decay. This behavior is characteristic of an ARMA(p,q) process rather than a pure AR(p) or MA(q) process. Therefore, we consider ARMA models for several combinations of *p* and *q*.

```{r}
pmax <- 5
qmax <- 5

AIC_mat <- matrix(NA, pmax, qmax)
BIC_mat <- matrix(NA, pmax, qmax)

for (p in 1:pmax) {
  for (q in 1:qmax) {
    
    fit <- arima(X4, order=c(p,0,q))
    
    AIC_mat[p,q] <- AIC(fit)
    BIC_mat[p,q] <- BIC(fit)
  }
}

knitr::kable(AIC_mat, caption="AIC values for ARMA(p,q)")
knitr::kable(BIC_mat, caption="BIC values for ARMA(p,q)")
```


```{r}
#| results: hold
which(AIC_mat == min(AIC_mat), arr.ind=TRUE)
which(BIC_mat == min(BIC_mat), arr.ind=TRUE)
which(AIC_mat==sort(AIC_mat)[2], arr.ind=TRUE)
which(BIC_mat==sort(BIC_mat)[2], arr.ind=TRUE)
```

The minimum AIC and BIC values correspond to the models ARMA(1,1) and ARMA(2,1).

# Exercice 2 - Maximum Likelihood parameter estimation

```{r}
fit11 <- arima(X4, order=c(1,0,1), method="ML")
fit21 <- arima(X4, order=c(2,0,1), method="ML")

fit11
fit21
```

# Exercice 3 - Model Equation

The equations of the two selected models are:
$$
ARMA(1,1): X_t= 0.6334 + 0.8816X_{t-1} + 0.8780\epsilon_{t-1} + \epsilon_t
$$
$$
ARMA(2,1) : X_t= 0.6277+ 0.8592X_{t-1} + 0.0246X_{t-2} + 0.8835\epsilon_{t-1} + \epsilon_t
$$

# Exercice 4 - Residual diagnostics

To determine whether the models are correctly specified, we analyze the residuals.
If the model adequately captures the dependence structure of the series, the
residuals should behave like **white noise**.

We therefore examine:
- the residual time series
- the ACF and PACF of the residuals
- Ljung–Box tests for remaining autocorrelation.

## Model 1 diagnostics

```{r}
#| resuts: hold
#| fig-width: 7
#| fig-height: 6

par(mfrow=c(3,1))
resid11=residuals(fit11)
plot(resid11)
acf(resid11)
pacf(resid11)
```

The residuals fluctuate randomly around zero and no clear pattern is visible.
The ACF and PACF show no significant spikes outside the confidence bands,
indicating the absence of remaining autocorrelation.

## Ljung- Box Test

```{r}
pval1 <- numeric(19)

for(i in 2:20){
  pval1[i-1] <- Box.test(resid11, lag=i,
                         type="Ljung",
                         fitdf=2)$p.value
}

plot(2:20, pval1,
     main="Ljung-Box p-values for ARMA(1,1)",
     xlab="Lag", ylab="p-value")

abline(h=0.05)
```

Most p-values are above the significance level (0.05), suggesting that
the residuals do not exhibit significant autocorrelation.
Therefore the ARMA(1,1) model appears to adequately capture the correlation
structure of the data.

## Model 2

```{r}
#| resuts: hold
#| fig-width: 7
#| fig-height: 6
par(mfrow=c(3,1))
resid21 <- residuals(fit21)
plot(resid21, main="Residuals of ARMA(2,1)")
acf(resid21, main="ACF of residuals")
pacf(resid21, main="PACF of residuals")
```

```{r}
pval2 <- numeric(19)

for(i in 2:20){
  pval2[i-1] <- Box.test(resid21, lag=i,
                         type="Ljung",
                         fitdf=3)$p.value
}

plot(2:20, pval2,
     main="Ljung-Box p-values for ARMA(2,1)",
     xlab="Lag", ylab="p-value")

abline(h=0.05)
```

Similarly, the residual diagnostics indicate that the ARMA(2,1) model
does not leave significant autocorrelation in the residuals. Thus this
model also appears to be correctly specified.

Between the two models, the **ARMA(1,1)** model is more parsimonious because
it contains fewer parameters than the *ARMA(2,1)* model.

Since both models adequately capture the correlation structure of the data, the ARMA(1,1) model is preferred.


# Exercice 5 - One-step-ahead prediction

To compare the predictive performance of the two models, we compute
one-step-ahead forecasts using a rolling window approach.

```{r}
OneAhead <- function(x, order, K = 0.9){

  n <- length(x)
  ntrain <- floor(K*n)

  preds <- numeric(n)

  for(i in ntrain:(n-1)){
    fit <- arima(x[1:i], order=order)
    preds[i+1] <- predict(fit, n.ahead=1)$pred
  }

  idx <- (ntrain+1):n
  pred <- preds[idx]
  errors <- x[idx] - pred

  list(pred=pred,
       errors=errors,
       mspe=mean(errors^2))
}
```

```{r}
#| results: hold
OAP11 <- OneAhead(X4, order=c(1,0,1))
OAP21 <- OneAhead(X4, order=c(2,0,1))

OAP11$mspe
OAP21$mspe
```

Thsi confirms our hypothesis made earlier. These two models are almost identical in terms of predicting the data.



For firther analysis, we plot the *actual* vs the *predicted*.

```{r}
n <- length(X4)
ntrain <- floor(0.9*n)

actual <- X4[(ntrain+1):n]

plot(actual,
     type="l",
     col="black",
     main="One-step-ahead predictions",
     ylab="Series value",
     xlab="Time")

lines(OAP11$pred, col="red")
lines(OAP21$pred, col="blue")

legend("topleft",
       legend=c("Actual","ARMA(1,1)","ARMA(2,1)"),
       col=c("black","red","blue"),
       lty=1)
```

Again, we observe the same thing, the one step ahead predictions of the ARMA(1,1) and the ARMA(2,1) are intertwined. They make the same predictions


# Exercice 6 
Both models appear to adequately capture the correlation structure of the data.
However, the ARMA(1,1) model is more parsimonious. Therefore, the **ARMA(1,1)** model is selected as the most appropriate model
for the series.

# Exercice 7 - Forecasting


```{r}
forecast <- predict(fit11, n.ahead = 10)

# last 50 observations
last50 <- tail(X4, 50)

plot(last50,
     xlim = c(time(last50)[1], time(last50)[50] + 10),
     ylim = range(c(last50, forecast$pred)),
     main = "10-step forecast with ARMA(1,1)", type="l",
     ylab = "Series",
     col = "black")

lines(seq(time(last50)[50] + deltat(X4),
          by = deltat(X4),
          length.out = 10),
      forecast$pred,
      col = "red")

legend("topleft",
       legend = c("Observed", "Forecast"),
       col = c("black","red"),
       lty = 1)

```

The figure shows the last 50 observations of the series together with the 10-step-ahead forecasts obtained from the selected ARMA(1,1) model.
